---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First clear out R and set the working directory
```{r}
rm(list=ls())
setwd("C:/_WORK/COURSERA/DataScience/ReproducibleResearch/Proj1/RepData_PeerAssessment1")
library("sqldf")
library("ggplot2")

```
Then read in the csv and change the type of the `date` column
```{r}
activity<-read.csv("activity.csv", header = TRUE, na.strings = "NA")

activity$date <- as.Date(activity$date , "%Y-%m-%d")
head(activity)
```


## What is mean total number of steps taken per day?
First subset the data
```{r}
activityDay<-sqldf("select date, sum(steps) as TotSteps from activity where steps<>'NA' group by date")
```
This is a histogram of the total steps
```{r}
p<-ggplot(data=activityDay, aes(activityDay$TotSteps)) + geom_histogram()+labs(title="Histogram of Total steps by day") +
  labs(x="Total Steps", y="Count") 
print(p)
```

The mean number of steps, rounded to 1 decimal place, is `r format(round(mean(activityDay$TotSteps), digits=1),nsmall=1)` and the median is `r median(activityDay$TotSteps)`

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
